---
title: "Cellwise Alpha Distance Analysis"
author: "Taher Alabbar"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
  pdf_document:
    toc: true
params:
  data_root: "data/panc8"
  all_batches:
    - "celseq"
    - "celseq2"
    - "fluidigmc1"
    - "indrop"
    - "smartseq2"
  primary_batches:
    - "indrop"
    - "smartseq2"
  include_all_batches: true
  alpha_label: "alpha"
  assay: "RNA"
  distance_layer: "data"
  feature_mode: "variable"
  n_variable_features: 2000
  min_cells: 3
  min_features: 200
  min_cells_per_type: 20
  max_cells_per_type: 500
  sample_per_group: 5000
  seed: 123
  write_outputs: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 7,
  dpi = 300,
  out.width = "100%"
)

library(Seurat)
library(Matrix)
library(dplyr)
library(ggplot2)
library(patchwork)

source(file.path("scripts", "cellwise_distance_utils.R"))

theme_set(theme_minimal(base_size = 11))
```

# Overview

This report compares alpha-centered cellwise distances across single-cell RNA-seq
batches. For each batch, alpha-to-other-cell-type distances are normalized by
the within-batch alpha-to-alpha median distance:

```text
relative distance =
  alpha-to-cell-type distance / within-batch alpha-to-alpha median
```

This normalization makes batch-level comparisons easier to interpret than raw
Euclidean distances alone. Values near 1 mean the comparison is close to the
within-alpha baseline; values above 1 indicate stronger separation from alpha.

# Analysis Settings

```{r analysis-settings}
set.seed(params$seed)

all_batches <- as.character(unlist(params$all_batches, use.names = FALSE))
primary_batches <- as.character(unlist(params$primary_batches, use.names = FALSE))

analysis_batches <- if (isTRUE(params$include_all_batches)) {
  all_batches
} else {
  primary_batches
}

if (!all(primary_batches %in% analysis_batches)) {
  stop("primary_batches must be included in the batches selected for analysis.")
}

data_root <- resolve_data_root(params$data_root)
dataset_dirs <- make_dataset_dirs(data_root, analysis_batches)
primary_batch_label <- make_batch_label(primary_batches)

data.frame(
  setting = c(
    "data_root",
    "analysis_batches",
    "primary_batches",
    "alpha_label",
    "distance_layer",
    "feature_mode",
    "max_cells_per_type"
  ),
  value = c(
    data_root,
    paste(analysis_batches, collapse = ", "),
    paste(primary_batches, collapse = ", "),
    params$alpha_label,
    params$distance_layer,
    params$feature_mode,
    params$max_cells_per_type
  )
)
```

# Load And Preprocess Data

```{r load-data}
seu <- load_10x_batches(
  dataset_dirs = dataset_dirs,
  min_cells = params$min_cells,
  min_features = params$min_features
)

seu
```

```{r preprocess-data}
seu <- NormalizeData(seu, verbose = FALSE)
seu <- FindVariableFeatures(
  seu,
  selection.method = "vst",
  nfeatures = params$n_variable_features,
  verbose = FALSE
)

if (identical(params$distance_layer, "scale.data")) {
  scale_features <- if (identical(params$feature_mode, "all")) {
    rownames(seu)
  } else {
    VariableFeatures(seu)
  }

  seu <- ScaleData(seu, features = scale_features, verbose = FALSE)
}
```

```{r cell-counts}
table(seu$batch)
table(seu$cell_type, seu$batch, useNA = "ifany")
```

```{r select-features}
expression_matrix <- get_expression_layer(
  seurat_object = seu,
  assay = params$assay,
  layer = params$distance_layer
)

distance_features <- select_distance_features(
  seurat_object = seu,
  expression_matrix = expression_matrix,
  feature_mode = params$feature_mode,
  n_features = params$n_variable_features
)

data.frame(
  distance_layer = params$distance_layer,
  feature_mode = params$feature_mode,
  n_features = length(distance_features)
)
```

# Alpha-Centered Distances

```{r calculate-distances}
alpha_results <- calculate_alpha_to_other_distances(
  seurat_object = seu,
  batches = analysis_batches,
  alpha_label = params$alpha_label,
  assay = params$assay,
  layer = params$distance_layer,
  features = distance_features,
  min_cells_per_type = params$min_cells_per_type,
  max_cells_per_type = params$max_cells_per_type,
  seed = params$seed
)

alpha_distances <- alpha_results$distances
alpha_baselines <- alpha_results$alpha_baselines
alpha_summary <- summarise_alpha_distances(alpha_distances)
```

```{r alpha-baselines}
alpha_baselines
```

```{r alpha-summary}
alpha_summary
```

# Primary Batch Comparison

```{r primary-comparison-data}
primary_distances <- filter_shared_cell_types(
  distance_table = alpha_distances,
  batches = primary_batches
)

primary_summary <- summarise_alpha_distances(primary_distances)

primary_sample <- sample_distance_table(
  distance_table = primary_distances,
  group_cols = c("batch", "comparison"),
  n_per_group = params$sample_per_group,
  seed = params$seed
)

primary_summary
```

```{r primary-relative-distributions}
plot_relative_distance_distribution(
  primary_sample,
  title = "Relative alpha-to-cell-type distance distributions",
  subtitle = primary_batch_label
)
```

```{r primary-alpha-beta}
alpha_beta_sample <- primary_sample |>
  filter(other_cell_type == "beta")

if (nrow(alpha_beta_sample) == 0) {
  message("No beta-cell comparison passed the filtering threshold.")
} else {
  plot_relative_distance_distribution(
    alpha_beta_sample,
    title = "Relative alpha-to-beta distance distributions",
    subtitle = primary_batch_label
  )
}
```

```{r primary-relative-means}
plot_mean_distance_summary(
  primary_summary,
  x_metric = "mean_relative_distance",
  title = "Mean relative alpha-to-cell-type distances",
  subtitle = primary_batch_label,
  x_label = "Mean distance relative to within-batch alpha baseline",
  add_baseline = TRUE
)
```

```{r primary-raw-distributions}
plot_raw_distance_distribution(
  primary_sample,
  title = "Raw alpha-to-cell-type distance distributions",
  subtitle = primary_batch_label
)
```

```{r raw-vs-relative}
plot_raw_vs_relative_summary(
  primary_distances,
  title = "Raw versus relative alpha-to-cell-type distances",
  subtitle = primary_batch_label
)
```

# All-Batch Overview

```{r all-batch-relative-differences}
relative_batch_differences <- pairwise_batch_differences(
  alpha_summary,
  metric_col = "mean_relative_distance"
)

relative_batch_differences
```

```{r all-batch-relative-heatmap}
plot_pairwise_difference_heatmap(
  relative_batch_differences,
  title = "Relative alpha-to-cell-type separation differences",
  subtitle = "Signed difference in mean relative distance for each batch pair",
  fill_label = "Signed mean\nrelative difference"
)
```

```{r all-batch-raw-differences}
raw_batch_differences <- pairwise_batch_differences(
  alpha_summary,
  metric_col = "mean_distance"
)

raw_batch_differences
```

```{r all-batch-raw-heatmap}
plot_pairwise_difference_heatmap(
  raw_batch_differences,
  title = "Raw alpha-to-cell-type separation differences",
  subtitle = "Signed difference in mean Euclidean distance for each batch pair",
  fill_label = "Signed mean\nraw difference"
)
```

# Export Tables

```{r export-tables}
if (isTRUE(params$write_outputs)) {
  dir.create("output", showWarnings = FALSE)

  write.csv(
    alpha_baselines,
    file = file.path("output", "alpha_baselines.csv"),
    row.names = FALSE
  )

  write.csv(
    alpha_summary,
    file = file.path("output", "alpha_distance_summary.csv"),
    row.names = FALSE
  )

  write.csv(
    relative_batch_differences,
    file = file.path("output", "relative_batch_differences.csv"),
    row.names = FALSE
  )

  write.csv(
    raw_batch_differences,
    file = file.path("output", "raw_batch_differences.csv"),
    row.names = FALSE
  )
}
```

# Session Info

```{r session-info}
sessionInfo()
```
