---
title: "Single-Cell Batch-Correction Evaluation"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
params:
  seurat_rds: !r NULL
  seurat_object_name: seu
  tenx_data_root: /Users/taheralabbar/Desktop/starter code and data/data/panc8
  tenx_batches: !r c("celseq", "celseq2", "fluidigmc1", "smartseq2", "indrop")
  batch_col: batch
  cell_type_col: cell_type
  assay: RNA
  reference_cell_type: alpha
  beta_cell_type: beta
  n_dims: 30
  n_variable_features: 2000
  scale_all_genes: true
  future_globals_max_size_gb: 8
  min_cells_per_group: 10
  max_cells_per_group: 500
  max_pairs_per_comparison: 10000
  seed: 123
  output_dir: batch_correction_evaluation_output
  selected_overlay_batch_pairs: !r NULL
  selected_overlay_cell_types: !r c("alpha", "beta")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.align = "center"
)
```

# Package Setup

This chunk is intentionally disabled by `run_package_install <- FALSE`. Only
set it to `TRUE` in a fresh R session if packages are missing. If you need to
upgrade already-loaded packages such as `dplyr`, restart R first and install
before loading `Seurat`, `tidyr`, `plotly`, or `sctransform`.

```{r install-packages, eval=FALSE}
run_package_install <- FALSE

if (run_package_install) {
  install_cran_if_missing <- function(packages) {
    missing_packages <- packages[
      !vapply(packages, requireNamespace, logical(1), quietly = TRUE)
    ]

    if (length(missing_packages) > 0) {
      install.packages(missing_packages)
    }
  }

  install_bioc_if_missing <- function(packages) {
    missing_packages <- packages[
      !vapply(packages, requireNamespace, logical(1), quietly = TRUE)
    ]

    if (length(missing_packages) > 0) {
      BiocManager::install(missing_packages, update = FALSE, ask = FALSE)
    }
  }

  install_cran_if_missing(c(
    "Seurat",
    "dplyr",
    "tidyr",
    "purrr",
    "ggplot2",
    "readr",
    "tibble",
    "future",
    "magrittr",
    "rlang",
    "Matrix",
    "remotes"
  ))

  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }

  install_bioc_if_missing(c(
    "SingleCellExperiment",
    "SummarizedExperiment",
    "S4Vectors",
    "batchelor",
    "scran",
    "scater",
    "BiocSingular",
    "BiocParallel"
  ))

  # Optional, useful if you also want Seurat wrapper methods such as RunFastMNN.
  if (!requireNamespace("SeuratWrappers", quietly = TRUE)) {
    remotes::install_github("satijalab/seurat-wrappers", upgrade = "never")
  }
}
```

```{r libraries}
required_packages <- c(
  "Seurat",
  "SeuratObject",
  "SingleCellExperiment",
  "SummarizedExperiment",
  "S4Vectors",
  "Matrix",
  "batchelor",
  "BiocSingular",
  "BiocParallel",
  "dplyr",
  "magrittr",
  "purrr",
  "ggplot2",
  "readr",
  "tibble",
  "future"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Missing required packages: ",
    paste(missing_packages, collapse = ", "),
    ". Restart R, install them in a fresh session, then knit again.",
    call. = FALSE
  )
}

# Use namespace-qualified local aliases instead of library(dplyr). This avoids
# dplyr unload/reload conflicts when Seurat dependencies already imported dplyr.
`%>%` <- magrittr::`%>%`
arrange <- dplyr::arrange
bind_rows <- dplyr::bind_rows
filter <- dplyr::filter
full_join <- dplyr::full_join
group_by <- dplyr::group_by
group_split <- dplyr::group_split
inner_join <- dplyr::inner_join
mutate <- dplyr::mutate
n <- dplyr::n
rename <- dplyr::rename
select <- dplyr::select
summarise <- dplyr::summarise
transmute <- dplyr::transmute

imap <- purrr::imap
map <- purrr::map
map_dfr <- purrr::map_dfr
tibble <- tibble::tibble

aes <- ggplot2::aes
element_blank <- ggplot2::element_blank
element_text <- ggplot2::element_text
facet_grid <- ggplot2::facet_grid
facet_wrap <- ggplot2::facet_wrap
geom_abline <- ggplot2::geom_abline
geom_density <- ggplot2::geom_density
geom_hline <- ggplot2::geom_hline
geom_point <- ggplot2::geom_point
geom_text <- ggplot2::geom_text
geom_tile <- ggplot2::geom_tile
geom_vline <- ggplot2::geom_vline
ggplot <- ggplot2::ggplot
labs <- ggplot2::labs
scale_fill_gradient2 <- ggplot2::scale_fill_gradient2
scale_fill_viridis_c <- ggplot2::scale_fill_viridis_c
theme <- ggplot2::theme
theme_minimal <- ggplot2::theme_minimal
vars <- ggplot2::vars
```

# Configuration

Adjust the parameters in the YAML header, or override these values below.

```{r config}
batch_col <- params$batch_col
cell_type_col <- params$cell_type_col
assay_name <- params$assay
reference_cell_type <- params$reference_cell_type
beta_cell_type <- params$beta_cell_type
n_dims <- params$n_dims
n_variable_features <- params$n_variable_features
scale_all_genes <- params$scale_all_genes
future_globals_max_size_gb <- params$future_globals_max_size_gb
min_cells_per_group <- params$min_cells_per_group
max_cells_per_group <- params$max_cells_per_group
max_pairs_per_comparison <- params$max_pairs_per_comparison
seed <- params$seed

options(future.globals.maxSize = future_globals_max_size_gb * 1024^3)
future::plan(future::sequential)

output_dir <- params$output_dir
figures_dir <- file.path(output_dir, "figures")
tables_dir <- file.path(output_dir, "tables")
dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tables_dir, recursive = TRUE, showWarnings = FALSE)
```

# Input Seurat Object

This report standardizes the input object to `seu`. Use one of these options:

- If your object is already loaded as `seu`, run the report directly.
- If your object has another name, set `seurat_object_name` in the YAML.
- If knitting in a fresh R session, set `seurat_rds` in the YAML to an `.rds`
  file created with `saveRDS(your_seurat_object, "path/to/file.rds")`.

```{r load-seurat-object}
seurat_rds_path <- params$seurat_rds
seurat_object_name <- params$seurat_object_name

# If running chunks interactively, you can temporarily override the YAML here:
# seurat_rds_path <- "/absolute/path/to/your_seurat_object.rds"
# seurat_object_name <- "your_object_name"

find_loaded_seurat_objects <- function() {
  object_names <- ls(envir = .GlobalEnv)
  object_names[
    vapply(
      object_names,
      function(object_name) {
        inherits(get(object_name, envir = .GlobalEnv), "Seurat")
      },
      logical(1)
    )
  ]
}

load_10x_batches_from_root <- function(data_root,
                                       batch_names,
                                       batch_col,
                                       cell_type_col) {
  if (is.null(data_root) || !dir.exists(data_root)) {
    stop(
      "The 10X data root does not exist: ",
      data_root,
      ". Set `tenx_data_root` to the folder containing the batch folders.",
      call. = FALSE
    )
  }

  dataset_dirs <- file.path(data_root, batch_names)
  names(dataset_dirs) <- batch_names

  missing_dirs <- dataset_dirs[!dir.exists(dataset_dirs)]

  if (length(missing_dirs) > 0) {
    stop(
      "These 10X batch directories do not exist:\n",
      paste(names(missing_dirs), missing_dirs, sep = ": ", collapse = "\n"),
      call. = FALSE
    )
  }

  objects <- lapply(names(dataset_dirs), function(batch_name) {
    counts <- Seurat::Read10X(data.dir = dataset_dirs[[batch_name]])

    obj <- Seurat::CreateSeuratObject(
      counts = counts,
      project = batch_name,
      min.cells = 3,
      min.features = 200
    )

    obj[[batch_col]] <- batch_name

    annotation_file <- file.path(dataset_dirs[[batch_name]], "annotations.csv")

    if (file.exists(annotation_file)) {
      annotations <- read.csv(
        annotation_file,
        row.names = "barcode",
        stringsAsFactors = FALSE
      )

      if (!cell_type_col %in% colnames(annotations)) {
        stop(
          "Annotation file for ",
          batch_name,
          " does not contain column `",
          cell_type_col,
          "`.",
          call. = FALSE
        )
      }

      obj[[cell_type_col]] <- annotations[colnames(obj), cell_type_col]
    } else {
      obj[[cell_type_col]] <- NA_character_
      warning("No annotations.csv found for batch: ", batch_name, call. = FALSE)
    }

    obj
  })

  merged <- if (length(objects) == 1) {
    objects[[1]]
  } else {
    merge(
      objects[[1]],
      y = objects[-1],
      add.cell.ids = names(dataset_dirs)
    )
  }

  if ("JoinLayers" %in% getNamespaceExports("SeuratObject")) {
    merged <- SeuratObject::JoinLayers(merged)
  }

  merged
}

if (exists("seu", inherits = TRUE) && inherits(seu, "Seurat")) {
  message("Using loaded Seurat object named `seu`.")
} else if (!is.null(seurat_rds_path) && file.exists(seurat_rds_path)) {
  seu <- readRDS(seurat_rds_path)
  message("Loaded Seurat object from: ", seurat_rds_path)
} else if (
  !is.null(seurat_object_name) &&
    exists(seurat_object_name, envir = .GlobalEnv, inherits = FALSE)
) {
  seu <- get(seurat_object_name, envir = .GlobalEnv)
  message("Using loaded Seurat object named `", seurat_object_name, "`.")
} else if (!is.null(params$tenx_data_root) && dir.exists(params$tenx_data_root)) {
  seu <- load_10x_batches_from_root(
    data_root = params$tenx_data_root,
    batch_names = params$tenx_batches,
    batch_col = batch_col,
    cell_type_col = cell_type_col
  )
  message("Loaded 10X batches from: ", params$tenx_data_root)
} else {
  loaded_seurat_objects <- find_loaded_seurat_objects()

  if (length(loaded_seurat_objects) == 1) {
    seu <- get(loaded_seurat_objects[[1]], envir = .GlobalEnv)
    message("Using the only loaded Seurat object: `", loaded_seurat_objects[[1]], "`.")
  } else {
    available_message <- if (length(loaded_seurat_objects) == 0) {
      "No loaded Seurat objects were detected in .GlobalEnv."
    } else {
      paste(
        "Loaded Seurat objects detected:",
        paste(loaded_seurat_objects, collapse = ", ")
      )
    }

    stop(
      "No input Seurat object was found. Load your object as `seu`, ",
      "set `seurat_object_name` to its object name, or set `seurat_rds` ",
      "to an existing .rds file, or set `tenx_data_root` to the folder ",
      "containing the batch directories. ",
      available_message,
      call. = FALSE
    )
  }
}

if (!inherits(seu, "Seurat")) {
  stop("`seu` must be a Seurat object.", call. = FALSE)
}
```

# Normalize And Select Variable Genes

This chunk normalizes `seu`, selects the top `n_variable_features` genes by VST
for PCA and MNN methods, and scales either all genes or only the selected
variable genes depending on `scale_all_genes`.

```{r qc-normalize}
set.seed(seed)
SeuratObject::DefaultAssay(seu) <- assay_name

seu <- Seurat::NormalizeData(seu, verbose = FALSE)
seu <- Seurat::FindVariableFeatures(
  seu,
  selection.method = "vst",
  nfeatures = n_variable_features,
  verbose = FALSE
)

pca_features <- SeuratObject::VariableFeatures(seu)

if (length(pca_features) == 0) {
  stop("No variable features were selected for PCA.", call. = FALSE)
}

features_to_scale <- if (isTRUE(scale_all_genes)) {
  rownames(seu)
} else {
  pca_features
}

seu <- Seurat::ScaleData(
  seu,
  features = features_to_scale,
  verbose = FALSE
)

message(
  "Normalized `seu`; selected ",
  length(pca_features),
  " variable genes for PCA/MNN; scaled ",
  length(features_to_scale),
  " genes."
)
```

# Helper Functions

```{r helper-functions}
safe_divide <- function(numerator, denominator) {
  out <- numerator / denominator
  bad <- !is.finite(out) | !is.finite(denominator) | denominator <= 0
  out[bad] <- NA_real_
  out
}

safe_abs_log <- function(x) {
  out <- abs(log(x))
  out[!is.finite(out) | x <= 0] <- NA_real_
  out
}

clean_labels <- function(x) {
  x <- as.character(x)
  x[is.na(x) | x == ""] <- NA_character_
  x
}

check_required_metadata <- function(seu, batch_col, cell_type_col, assay_name) {
  metadata <- seu[[]]
  missing_cols <- setdiff(c(batch_col, cell_type_col), colnames(metadata))

  if (length(missing_cols) > 0) {
    stop(
      "Missing metadata columns: ",
      paste(missing_cols, collapse = ", "),
      call. = FALSE
    )
  }

  if (!assay_name %in% names(seu@assays)) {
    stop("Assay `", assay_name, "` was not found in `seu`.", call. = FALSE)
  }

  metadata[[batch_col]] <- clean_labels(metadata[[batch_col]])
  metadata[[cell_type_col]] <- clean_labels(metadata[[cell_type_col]])

  if (sum(!is.na(metadata[[batch_col]])) < 2) {
    stop("The batch column has fewer than two non-missing cells.", call. = FALSE)
  }

  if (sum(!is.na(metadata[[cell_type_col]])) < 2) {
    stop("The cell-type column has fewer than two non-missing cells.", call. = FALSE)
  }

  metadata
}

sample_vector <- function(x, max_n) {
  x <- as.character(x)

  if (is.null(max_n) || is.na(max_n) || length(x) <= max_n) {
    return(x)
  }

  sample(x, max_n)
}

effective_rank <- function(requested_rank, n_cells, n_features) {
  max(1L, min(as.integer(requested_rank), as.integer(n_cells) - 1L, as.integer(n_features)))
}

get_seurat_layer <- function(seu, assay_name, layer_name) {
  tryCatch(
    SeuratObject::LayerData(seu, assay = assay_name, layer = layer_name),
    error = function(e) {
      SeuratObject::GetAssayData(seu, assay = assay_name, slot = layer_name)
    }
  )
}

make_sce_from_seurat <- function(seu, assay_name) {
  logcounts_mat <- get_seurat_layer(seu, assay_name, "data")

  if (nrow(logcounts_mat) == 0 || ncol(logcounts_mat) == 0) {
    stop(
      "No normalized data layer was found. Run NormalizeData() before MNN methods.",
      call. = FALSE
    )
  }

  assays <- list(logcounts = logcounts_mat)

  counts_mat <- tryCatch(
    get_seurat_layer(seu, assay_name, "counts"),
    error = function(e) NULL
  )

  if (!is.null(counts_mat) && identical(dim(counts_mat), dim(logcounts_mat))) {
    assays$counts <- counts_mat[rownames(logcounts_mat), colnames(logcounts_mat), drop = FALSE]
  }

  sce <- SingleCellExperiment::SingleCellExperiment(assays = assays)
  SummarizedExperiment::colData(sce) <- S4Vectors::DataFrame(
    seu[[]][colnames(sce), , drop = FALSE]
  )

  sce
}

align_embedding <- function(embedding, metadata, method_name) {
  embedding <- as.matrix(embedding)

  if (is.null(rownames(embedding))) {
    stop("Embedding for ", method_name, " has no cell rownames.", call. = FALSE)
  }

  if (anyDuplicated(rownames(embedding)) > 0) {
    stop("Embedding for ", method_name, " has duplicated cell rownames.", call. = FALSE)
  }

  missing_cells <- setdiff(rownames(metadata), rownames(embedding))

  if (length(missing_cells) > 0) {
    stop(
      "Embedding for ",
      method_name,
      " is missing ",
      length(missing_cells),
      " metadata cells. First missing cell: ",
      missing_cells[[1]],
      call. = FALSE
    )
  }

  embedding[rownames(metadata), , drop = FALSE]
}

run_uncorrected_pca_embedding <- function(seu,
                                          assay_name,
                                          n_dims,
                                          n_variable_features,
                                          seed) {
  set.seed(seed)
  SeuratObject::DefaultAssay(seu) <- assay_name

  features <- SeuratObject::VariableFeatures(seu)

  if (length(features) == 0) {
    seu <- Seurat::NormalizeData(seu, verbose = FALSE)
    seu <- Seurat::FindVariableFeatures(
      seu,
      selection.method = "vst",
      nfeatures = n_variable_features,
      verbose = FALSE
    )
    features <- SeuratObject::VariableFeatures(seu)
  }

  features <- head(features, n_variable_features)
  rank <- effective_rank(n_dims, ncol(seu), length(features))

  scale_data <- tryCatch(
    get_seurat_layer(seu, assay_name, "scale.data"),
    error = function(e) NULL
  )

  missing_scaled_features <- if (is.null(scale_data) || nrow(scale_data) == 0) {
    features
  } else {
    setdiff(features, rownames(scale_data))
  }

  if (length(missing_scaled_features) > 0) {
    seu <- Seurat::ScaleData(seu, features = features, verbose = FALSE)
  }

  seu <- Seurat::RunPCA(
    seu,
    features = features,
    npcs = rank,
    reduction.name = "pca_uncorrected",
    seed.use = seed,
    verbose = FALSE
  )

  embedding <- SeuratObject::Embeddings(seu, reduction = "pca_uncorrected")
  embedding <- embedding[, seq_len(rank), drop = FALSE]

  list(
    object = seu,
    embedding = embedding,
    features = features
  )
}

run_fastmnn_embedding <- function(seu,
                                  metadata,
                                  batch_col,
                                  assay_name,
                                  features,
                                  n_dims,
                                  seed,
                                  k = 20) {
  set.seed(seed)
  sce <- make_sce_from_seurat(seu, assay_name)
  batch <- factor(metadata[colnames(sce), batch_col])
  features <- intersect(features, rownames(sce))
  rank <- effective_rank(n_dims, ncol(sce), length(features))

  fastmnn_out <- batchelor::fastMNN(
    sce,
    batch = batch,
    subset.row = features,
    d = rank,
    k = k,
    assay.type = "logcounts",
    BPPARAM = BiocParallel::SerialParam()
  )

  embedding <- SingleCellExperiment::reducedDim(fastmnn_out, "corrected")
  rownames(embedding) <- colnames(fastmnn_out)
  colnames(embedding) <- paste0("FastMNN_", seq_len(ncol(embedding)))
  embedding
}

run_mnn_embedding <- function(seu,
                              metadata,
                              batch_col,
                              assay_name,
                              features,
                              n_dims,
                              seed,
                              k = 20,
                              sigma = 0.1) {
  set.seed(seed)
  sce <- make_sce_from_seurat(seu, assay_name)
  batch <- factor(metadata[colnames(sce), batch_col])
  features <- intersect(features, rownames(sce))

  mnn_out <- batchelor::mnnCorrect(
    sce,
    batch = batch,
    subset.row = features,
    k = k,
    sigma = sigma,
    assay.type = "logcounts",
    correct.all = FALSE,
    BPPARAM = BiocParallel::SerialParam()
  )

  corrected <- SummarizedExperiment::assay(mnn_out, "corrected")
  rank <- effective_rank(n_dims, ncol(corrected), nrow(corrected))

  pca <- BiocSingular::runPCA(
    t(corrected),
    rank = rank,
    center = TRUE,
    scale = FALSE,
    BSPARAM = BiocSingular::IrlbaParam(),
    BPPARAM = BiocParallel::SerialParam()
  )

  embedding <- pca$x
  rownames(embedding) <- colnames(corrected)
  colnames(embedding) <- paste0("MNN_PC_", seq_len(ncol(embedding)))
  embedding
}

run_seurat_cca_embedding <- function(seu,
                                     batch_col,
                                     assay_name,
                                     n_dims,
                                     n_variable_features,
                                     seed) {
  set.seed(seed)
  future::plan(future::sequential)
  SeuratObject::DefaultAssay(seu) <- assay_name

  object_list <- Seurat::SplitObject(seu, split.by = batch_col)
  object_list <- object_list[vapply(object_list, ncol, numeric(1)) > 1]

  if (length(object_list) < 2) {
    stop("Seurat CCA requires at least two batches with more than one cell.", call. = FALSE)
  }

  object_list <- lapply(object_list, function(obj) {
    SeuratObject::DefaultAssay(obj) <- assay_name
    obj <- Seurat::NormalizeData(obj, verbose = FALSE)
    Seurat::FindVariableFeatures(
      obj,
      selection.method = "vst",
      nfeatures = n_variable_features,
      verbose = FALSE
    )
  })

  features <- Seurat::SelectIntegrationFeatures(
    object.list = object_list,
    nfeatures = n_variable_features
  )

  min_batch_cells <- min(vapply(object_list, ncol, numeric(1)))
  rank <- effective_rank(n_dims, min_batch_cells, length(features))
  dims <- seq_len(rank)

  object_list <- lapply(object_list, function(obj) {
    Seurat::ScaleData(obj, features = features, verbose = FALSE)
  })

  k_anchor <- min(5L, min_batch_cells - 1L)
  k_filter <- min(200L, min_batch_cells - 1L)
  k_score <- min(30L, min_batch_cells - 1L)
  k_weight <- min(100L, min_batch_cells - 1L)

  anchors <- Seurat::FindIntegrationAnchors(
    object.list = object_list,
    anchor.features = features,
    reduction = "cca",
    dims = dims,
    k.anchor = k_anchor,
    k.filter = k_filter,
    k.score = k_score,
    verbose = FALSE
  )

  integrated <- Seurat::IntegrateData(
    anchorset = anchors,
    new.assay.name = "integrated",
    dims = dims,
    k.weight = k_weight,
    verbose = FALSE
  )

  SeuratObject::DefaultAssay(integrated) <- "integrated"
  integrated <- Seurat::ScaleData(integrated, verbose = FALSE)
  integrated <- Seurat::RunPCA(
    integrated,
    npcs = rank,
    reduction.name = "integrated_pca",
    seed.use = seed,
    verbose = FALSE
  )

  embedding <- SeuratObject::Embeddings(integrated, reduction = "integrated_pca")
  embedding <- embedding[, seq_len(rank), drop = FALSE]
  colnames(embedding) <- paste0("SeuratCCA_PC_", seq_len(ncol(embedding)))

  list(
    object = integrated,
    embedding = embedding,
    features = features
  )
}

make_between_pairs <- function(cells_x, cells_y, max_pairs) {
  n_pairs_total <- length(cells_x) * length(cells_y)
  n_to_sample <- min(max_pairs, n_pairs_total)

  if (n_pairs_total <= max_pairs) {
    grid <- expand.grid(
      cell_x = cells_x,
      cell_y = cells_y,
      KEEP.OUT.ATTRS = FALSE,
      stringsAsFactors = FALSE
    )
    return(grid)
  }

  tibble(
    cell_x = sample(cells_x, n_to_sample, replace = TRUE),
    cell_y = sample(cells_y, n_to_sample, replace = TRUE)
  )
}

make_within_pairs <- function(cells, max_pairs) {
  if (length(cells) < 2) {
    return(tibble(cell_x = character(), cell_y = character()))
  }

  n_pairs_total <- length(cells) * (length(cells) - 1) / 2
  n_to_sample <- min(max_pairs, n_pairs_total)

  if (n_pairs_total <= max_pairs) {
    pair_matrix <- utils::combn(cells, 2)
    return(tibble(cell_x = pair_matrix[1, ], cell_y = pair_matrix[2, ]))
  }

  cell_x <- character()
  cell_y <- character()

  while (length(cell_x) < n_to_sample) {
    remaining <- n_to_sample - length(cell_x)
    candidate_x <- sample(cells, remaining * 2, replace = TRUE)
    candidate_y <- sample(cells, remaining * 2, replace = TRUE)
    keep <- candidate_x != candidate_y

    cell_x <- c(cell_x, candidate_x[keep])
    cell_y <- c(cell_y, candidate_y[keep])
  }

  tibble(
    cell_x = cell_x[seq_len(n_to_sample)],
    cell_y = cell_y[seq_len(n_to_sample)]
  )
}

paired_euclidean <- function(embedding, cell_x, cell_y) {
  x <- embedding[cell_x, , drop = FALSE]
  y <- embedding[cell_y, , drop = FALSE]
  sqrt(rowSums((x - y)^2))
}

sample_between_distances <- function(embedding, cells_x, cells_y, max_pairs) {
  pairs <- make_between_pairs(cells_x, cells_y, max_pairs)
  pairs$distance <- paired_euclidean(embedding, pairs$cell_x, pairs$cell_y)
  pairs
}

sample_within_distances <- function(embedding, cells, max_pairs) {
  pairs <- make_within_pairs(cells, max_pairs)

  if (nrow(pairs) == 0) {
    pairs$distance <- numeric()
    return(pairs)
  }

  pairs$distance <- paired_euclidean(embedding, pairs$cell_x, pairs$cell_y)
  pairs
}

cells_for_group <- function(metadata,
                            batch_col,
                            cell_type_col,
                            batch,
                            cell_type,
                            max_cells_per_group) {
  cells <- rownames(metadata)[
    metadata[[batch_col]] == batch &
      metadata[[cell_type_col]] == cell_type &
      !is.na(metadata[[batch_col]]) &
      !is.na(metadata[[cell_type_col]])
  ]

  sample_vector(cells, max_cells_per_group)
}

compute_technical_metrics_for_embedding <- function(embedding,
                                                    metadata,
                                                    batch_col,
                                                    cell_type_col,
                                                    method,
                                                    min_cells_per_group,
                                                    max_cells_per_group,
                                                    max_pairs_per_comparison,
                                                    seed) {
  set.seed(seed)
  metadata <- metadata[rownames(embedding), , drop = FALSE]
  metadata[[batch_col]] <- clean_labels(metadata[[batch_col]])
  metadata[[cell_type_col]] <- clean_labels(metadata[[cell_type_col]])

  batches <- sort(unique(stats::na.omit(metadata[[batch_col]])))
  cell_types <- sort(unique(stats::na.omit(metadata[[cell_type_col]])))
  batch_pairs <- utils::combn(batches, 2, simplify = FALSE)

  summary_rows <- list()
  distance_rows <- list()
  row_index <- 1L
  distance_index <- 1L

  for (batch_pair in batch_pairs) {
    batch_1 <- batch_pair[[1]]
    batch_2 <- batch_pair[[2]]
    batch_pair_label <- paste(batch_1, batch_2, sep = " vs ")

    for (cell_type in cell_types) {
      cells_1 <- cells_for_group(
        metadata,
        batch_col,
        cell_type_col,
        batch_1,
        cell_type,
        max_cells_per_group
      )
      cells_2 <- cells_for_group(
        metadata,
        batch_col,
        cell_type_col,
        batch_2,
        cell_type,
        max_cells_per_group
      )

      if (length(cells_1) < min_cells_per_group || length(cells_2) < min_cells_per_group) {
        message(
          "Skipping technical metric for ",
          method,
          ", ",
          cell_type,
          ", ",
          batch_pair_label,
          ": too few cells."
        )
        next
      }

      cross_distances <- sample_between_distances(
        embedding,
        cells_1,
        cells_2,
        max_pairs_per_comparison
      )
      within_1 <- sample_within_distances(embedding, cells_1, max_pairs_per_comparison)
      within_2 <- sample_within_distances(embedding, cells_2, max_pairs_per_comparison)

      if (nrow(within_1) == 0 || nrow(within_2) == 0) {
        message(
          "Skipping technical metric for ",
          method,
          ", ",
          cell_type,
          ", ",
          batch_pair_label,
          ": no within-batch distances."
        )
        next
      }

      within_baseline <- median(c(within_1$distance, within_2$distance), na.rm = TRUE)
      median_cross <- median(cross_distances$distance, na.rm = TRUE)
      r_tech <- safe_divide(median_cross, within_baseline)

      summary_rows[[row_index]] <- tibble(
        method = method,
        batch_1 = batch_1,
        batch_2 = batch_2,
        batch_pair = batch_pair_label,
        cell_type = cell_type,
        n_cells_batch_1 = length(cells_1),
        n_cells_batch_2 = length(cells_2),
        n_cross_distances = nrow(cross_distances),
        n_within_distances_batch_1 = nrow(within_1),
        n_within_distances_batch_2 = nrow(within_2),
        median_cross_batch_distance = median_cross,
        within_baseline = within_baseline,
        R_tech = r_tech,
        tech_deviation = safe_abs_log(r_tech)
      )
      row_index <- row_index + 1L

      distance_rows[[distance_index]] <- bind_rows(
        cross_distances %>%
          transmute(distance = distance) %>%
          mutate(distance_type = "Cross batch", source_batch = NA_character_),
        within_1 %>%
          transmute(distance = distance) %>%
          mutate(distance_type = "Within batch", source_batch = batch_1),
        within_2 %>%
          transmute(distance = distance) %>%
          mutate(distance_type = "Within batch", source_batch = batch_2)
      ) %>%
        mutate(
          method = method,
          batch_1 = batch_1,
          batch_2 = batch_2,
          batch_pair = batch_pair_label,
          cell_type = cell_type,
          within_baseline = within_baseline,
          normalized_distance = safe_divide(distance, within_baseline),
          .before = 1
        )
      distance_index <- distance_index + 1L
    }
  }

  list(
    summary = bind_rows(summary_rows),
    distances = bind_rows(distance_rows)
  )
}

compute_biological_metrics_for_embedding <- function(embedding,
                                                     metadata,
                                                     batch_col,
                                                     cell_type_col,
                                                     method,
                                                     reference_cell_type,
                                                     min_cells_per_group,
                                                     max_cells_per_group,
                                                     max_pairs_per_comparison,
                                                     seed) {
  set.seed(seed)
  metadata <- metadata[rownames(embedding), , drop = FALSE]
  metadata[[batch_col]] <- clean_labels(metadata[[batch_col]])
  metadata[[cell_type_col]] <- clean_labels(metadata[[cell_type_col]])

  batches <- sort(unique(stats::na.omit(metadata[[batch_col]])))

  summary_rows <- list()
  distance_rows <- list()
  row_index <- 1L
  distance_index <- 1L

  for (batch in batches) {
    batch_metadata <- metadata[
      metadata[[batch_col]] == batch &
        !is.na(metadata[[cell_type_col]]),
      ,
      drop = FALSE
    ]

    alpha_cells <- rownames(batch_metadata)[
      batch_metadata[[cell_type_col]] == reference_cell_type
    ]
    alpha_cells <- sample_vector(alpha_cells, max_cells_per_group)

    if (length(alpha_cells) < max(2L, min_cells_per_group)) {
      message(
        "Skipping biological metric for ",
        method,
        ", ",
        batch,
        ": fewer than ",
        max(2L, min_cells_per_group),
        " ",
        reference_cell_type,
        " cells."
      )
      next
    }

    alpha_within <- sample_within_distances(
      embedding,
      alpha_cells,
      max_pairs_per_comparison
    )
    alpha_baseline <- median(alpha_within$distance, na.rm = TRUE)

    if (!is.finite(alpha_baseline) || alpha_baseline <= 0) {
      message(
        "Skipping biological metric for ",
        method,
        ", ",
        batch,
        ": invalid alpha baseline."
      )
      next
    }

    target_cell_types <- sort(unique(stats::na.omit(batch_metadata[[cell_type_col]])))

    for (target_cell_type in target_cell_types) {
      target_cells <- rownames(batch_metadata)[
        batch_metadata[[cell_type_col]] == target_cell_type
      ]

      if (target_cell_type == reference_cell_type) {
        distances <- alpha_within
        n_target_cells <- length(alpha_cells)
      } else {
        target_cells <- sample_vector(target_cells, max_cells_per_group)

        if (length(target_cells) < min_cells_per_group) {
          message(
            "Skipping biological metric for ",
            method,
            ", ",
            batch,
            ", ",
            reference_cell_type,
            " vs ",
            target_cell_type,
            ": too few target cells."
          )
          next
        }

        distances <- sample_between_distances(
          embedding,
          alpha_cells,
          target_cells,
          max_pairs_per_comparison
        )
        n_target_cells <- length(target_cells)
      }

      distances <- distances %>%
        mutate(relative_distance = safe_divide(distance, alpha_baseline))

      summary_rows[[row_index]] <- tibble(
        method = method,
        batch = batch,
        reference_cell_type = reference_cell_type,
        target_cell_type = target_cell_type,
        n_reference_cells = length(alpha_cells),
        n_target_cells = n_target_cells,
        n_pairwise_distances = nrow(distances),
        alpha_baseline = alpha_baseline,
        mean_relative_separation = mean(distances$relative_distance, na.rm = TRUE),
        median_relative_separation = median(distances$relative_distance, na.rm = TRUE)
      )
      row_index <- row_index + 1L

      distance_rows[[distance_index]] <- distances %>%
        transmute(
          method = method,
          batch = batch,
          reference_cell_type = reference_cell_type,
          target_cell_type = target_cell_type,
          alpha_baseline = alpha_baseline,
          distance = distance,
          relative_distance = relative_distance
        )
      distance_index <- distance_index + 1L
    }
  }

  list(
    summary = bind_rows(summary_rows),
    distances = bind_rows(distance_rows)
  )
}

compute_biological_geometry_delta <- function(biological_summary) {
  split_groups <- biological_summary %>%
    group_by(method, target_cell_type) %>%
    group_split()

  map_dfr(split_groups, function(group_df) {
    batches <- sort(unique(group_df$batch))

    if (length(batches) < 2) {
      return(tibble())
    }

    map_dfr(utils::combn(batches, 2, simplify = FALSE), function(batch_pair) {
      batch_1 <- batch_pair[[1]]
      batch_2 <- batch_pair[[2]]

      value_1 <- group_df$mean_relative_separation[group_df$batch == batch_1][[1]]
      value_2 <- group_df$mean_relative_separation[group_df$batch == batch_2][[1]]

      tibble(
        method = group_df$method[[1]],
        target_cell_type = group_df$target_cell_type[[1]],
        batch_1 = batch_1,
        batch_2 = batch_2,
        batch_pair = paste(batch_1, batch_2, sep = " vs "),
        mean_relative_batch_1 = value_1,
        mean_relative_batch_2 = value_2,
        signed_delta = value_1 - value_2
      )
    })
  })
}

make_technical_scatter_data <- function(technical_summary,
                                        baseline_method = "Uncorrected_PCA") {
  baseline <- technical_summary %>%
    filter(method == baseline_method) %>%
    select(batch_pair, cell_type, R_tech_before = R_tech)

  technical_summary %>%
    filter(method != baseline_method) %>%
    inner_join(baseline, by = c("batch_pair", "cell_type")) %>%
    rename(R_tech_after = R_tech)
}

make_biological_scatter_data <- function(biological_summary,
                                         baseline_method = "Uncorrected_PCA") {
  baseline <- biological_summary %>%
    filter(method == baseline_method) %>%
    select(
      batch,
      target_cell_type,
      mean_relative_before = mean_relative_separation
    )

  biological_summary %>%
    filter(method != baseline_method) %>%
    inner_join(baseline, by = c("batch", "target_cell_type")) %>%
    rename(mean_relative_after = mean_relative_separation)
}

make_final_method_summary <- function(technical_summary,
                                      biological_summary,
                                      biological_delta,
                                      beta_cell_type,
                                      baseline_method = "Uncorrected_PCA") {
  technical_scores <- technical_summary %>%
    group_by(method) %>%
    summarise(
      technical_score = median(tech_deviation, na.rm = TRUE),
      median_R_tech = median(R_tech, na.rm = TRUE),
      n_technical_comparisons = n(),
      .groups = "drop"
    )

  baseline_bio <- biological_summary %>%
    filter(method == baseline_method) %>%
    select(
      batch,
      target_cell_type,
      mean_relative_before = mean_relative_separation
    )

  biological_preservation_scores <- biological_summary %>%
    inner_join(baseline_bio, by = c("batch", "target_cell_type")) %>%
    mutate(
      relative_to_uncorrected = safe_divide(
        mean_relative_separation,
        mean_relative_before
      ),
      biological_preservation_error = safe_abs_log(relative_to_uncorrected)
    ) %>%
    group_by(method) %>%
    summarise(
      biological_preservation_score = median(
        biological_preservation_error,
        na.rm = TRUE
      ),
      n_biological_comparisons = n(),
      .groups = "drop"
    )

  batch_agreement_scores <- biological_delta %>%
    group_by(method) %>%
    summarise(
      biological_batch_agreement_score = median(abs(signed_delta), na.rm = TRUE),
      n_biological_batch_pairs = n(),
      .groups = "drop"
    )

  alpha_beta_scores <- biological_summary %>%
    filter(target_cell_type == beta_cell_type) %>%
    group_by(method) %>%
    summarise(
      alpha_beta_collapse_risk = median(mean_relative_separation, na.rm = TRUE),
      n_beta_batches = n(),
      .groups = "drop"
    )

  technical_scores %>%
    full_join(biological_preservation_scores, by = "method") %>%
    full_join(batch_agreement_scores, by = "method") %>%
    full_join(alpha_beta_scores, by = "method") %>%
    arrange(technical_score)
}

filter_overlay_distances <- function(distance_table,
                                     selected_batch_pairs = NULL,
                                     selected_cell_types = NULL) {
  out <- distance_table

  if (!is.null(selected_batch_pairs)) {
    out <- out %>% filter(batch_pair %in% selected_batch_pairs)
  }

  if (!is.null(selected_cell_types)) {
    out <- out %>% filter(cell_type %in% selected_cell_types)
  }

  out
}

save_plot <- function(plot, filename, width = 10, height = 7, dpi = 300) {
  ggplot2::ggsave(
    filename = file.path(figures_dir, filename),
    plot = plot,
    width = width,
    height = height,
    dpi = dpi
  )
}
```

# Plot Functions

```{r plot-functions}
plot_technical_ratio_heatmap <- function(technical_summary) {
  ggplot(
    technical_summary,
    aes(x = batch_pair, y = cell_type, fill = R_tech)
  ) +
    geom_tile(color = "white", linewidth = 0.35) +
    geom_text(aes(label = round(R_tech, 2)), size = 2.8) +
    facet_wrap(vars(method)) +
    scale_fill_gradient2(
      low = "#2B6CB0",
      mid = "white",
      high = "#C2410C",
      midpoint = 1,
      na.value = "grey90"
    ) +
    labs(
      title = "Technical ratio by cell type and batch pair",
      subtitle = "R_tech = median cross-batch same-cell-type distance / within-batch baseline",
      x = "Batch pair",
      y = "Cell type",
      fill = "R_tech"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}

plot_technical_deviation_heatmap <- function(technical_summary) {
  ggplot(
    technical_summary,
    aes(x = batch_pair, y = cell_type, fill = tech_deviation)
  ) +
    geom_tile(color = "white", linewidth = 0.35) +
    geom_text(aes(label = round(tech_deviation, 2)), size = 2.8) +
    facet_wrap(vars(method)) +
    scale_fill_viridis_c(option = "C", direction = -1, na.value = "grey90") +
    ggplot2::coord_fixed(ratio = 1) +
    labs(
      title = "Technical deviation from ideal mixing",
      subtitle = "Lower abs(log(R_tech)) is better",
      x = "Batch pair",
      y = "Cell type",
      fill = "abs(log(R_tech))"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}

plot_technical_scatter <- function(technical_scatter) {
  ggplot(
    technical_scatter,
    aes(x = R_tech_before, y = R_tech_after, color = cell_type)
  ) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey45") +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey25") +
    geom_point(size = 2.2, alpha = 0.85) +
    facet_wrap(vars(method)) +
    labs(
      title = "Before-vs-after technical ratio",
      subtitle = "Each point is one cell type and batch pair",
      x = "Uncorrected PCA R_tech",
      y = "Corrected R_tech",
      color = "Cell type"
    ) +
    theme_minimal(base_size = 11)
}

plot_cross_vs_within_overlay <- function(distance_table) {
  ggplot(
    distance_table,
    aes(x = normalized_distance, fill = distance_type, color = distance_type)
  ) +
    geom_density(alpha = 0.25, linewidth = 0.5, na.rm = TRUE) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey35") +
    facet_grid(
      cell_type + batch_pair ~ method,
      scales = "free_y",
      switch = "y"
    ) +
    labs(
      title = "Cross-batch and within-batch normalized distance distributions",
      subtitle = "Distances are divided by the within-batch same-cell-type baseline inside each representation",
      x = "Normalized distance",
      y = "Density",
      fill = "Distance type",
      color = "Distance type"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "right",
      strip.placement = "outside",
      strip.text.x = element_text(size = 13),
      strip.text.y.left = element_text(angle = 0, size = 9, hjust = 1),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 13),
      plot.title = element_text(size = 18),
      plot.subtitle = element_text(size = 13),
      panel.spacing.x = grid::unit(1.1, "lines"),
      panel.spacing.y = grid::unit(0.35, "lines")
    )
}

plot_alpha_beta_distribution <- function(biological_distances, beta_cell_type) {
  biological_distances %>%
    filter(target_cell_type == beta_cell_type) %>%
    ggplot(aes(x = relative_distance, color = method, fill = method)) +
    geom_density(alpha = 0.18, linewidth = 0.55, na.rm = TRUE) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey35") +
    facet_wrap(vars(batch), scales = "free_y") +
    labs(
      title = paste("Relative alpha-to-", beta_cell_type, " distance distributions", sep = ""),
      subtitle = "Values mostly above 1 indicate separation from the alpha baseline",
      x = "Pairwise alpha-to-target distance / alpha-within-batch baseline",
      y = "Density",
      color = "Method",
      fill = "Method"
    ) +
    theme_minimal(base_size = 11)
}

plot_mean_relative_heatmap <- function(biological_summary) {
  ggplot(
    biological_summary,
    aes(x = batch, y = target_cell_type, fill = mean_relative_separation)
  ) +
    geom_tile(color = "white", linewidth = 0.35) +
    geom_text(aes(label = round(mean_relative_separation, 2)), size = 2.8) +
    facet_wrap(vars(method)) +
    scale_fill_viridis_c(option = "D", na.value = "grey90") +
    labs(
      title = "Mean alpha-centered relative separation",
      subtitle = "Mean of pairwise relative distances, not a ratio of medians",
      x = "Batch",
      y = "Target cell type",
      fill = "Mean relative\nseparation"
    ) +
    theme_minimal(base_size = 11) +
    theme(panel.grid = element_blank())
}

plot_biological_scatter <- function(biological_scatter) {
  ggplot(
    biological_scatter,
    aes(
      x = mean_relative_before,
      y = mean_relative_after,
      color = target_cell_type
    )
  ) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey45") +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey25") +
    geom_point(size = 2.2, alpha = 0.85) +
    facet_wrap(vars(method)) +
    labs(
      title = "Biological preservation relative to uncorrected PCA",
      subtitle = "Movement toward 1 can indicate biological collapse",
      x = "Uncorrected PCA mean relative separation",
      y = "Corrected mean relative separation",
      color = "Target cell type"
    ) +
    theme_minimal(base_size = 11)
}

plot_signed_delta_heatmap <- function(biological_delta) {
  ggplot(
    biological_delta,
    aes(x = batch_pair, y = target_cell_type, fill = signed_delta)
  ) +
    geom_tile(color = "white", linewidth = 0.35) +
    geom_text(aes(label = round(signed_delta, 2)), size = 2.8) +
    facet_wrap(vars(method)) +
    scale_fill_gradient2(
      low = "#2B6CB0",
      mid = "white",
      high = "#C2410C",
      midpoint = 0,
      na.value = "grey90"
    ) +
    labs(
      title = "Signed biological geometry difference across batches",
      subtitle = "Delta = mean relative separation in batch 1 minus batch 2",
      x = "Batch pair",
      y = "Target cell type",
      fill = "Signed delta"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```

# Build Representations

All distance comparisons below are made inside a representation and then
normalized inside that same representation. The code does not compare raw
original-space distances directly against corrected embedding distances.

```{r build-representations}
metadata <- check_required_metadata(seu, batch_col, cell_type_col, assay_name)

pca_result <- run_uncorrected_pca_embedding(
  seu = seu,
  assay_name = assay_name,
  n_dims = n_dims,
  n_variable_features = n_variable_features,
  seed = seed
)

metadata <- check_required_metadata(pca_result$object, batch_col, cell_type_col, assay_name)
metadata <- metadata[colnames(pca_result$object), , drop = FALSE]

fastmnn_embedding <- run_fastmnn_embedding(
  seu = pca_result$object,
  metadata = metadata,
  batch_col = batch_col,
  assay_name = assay_name,
  features = pca_result$features,
  n_dims = n_dims,
  seed = seed + 1
)

mnn_embedding <- run_mnn_embedding(
  seu = pca_result$object,
  metadata = metadata,
  batch_col = batch_col,
  assay_name = assay_name,
  features = pca_result$features,
  n_dims = n_dims,
  seed = seed + 2
)

cca_result <- run_seurat_cca_embedding(
  seu = seu,
  batch_col = batch_col,
  assay_name = assay_name,
  n_dims = n_dims,
  n_variable_features = n_variable_features,
  seed = seed + 3
)

embeddings <- list(
  Uncorrected_PCA = pca_result$embedding,
  FastMNN = fastmnn_embedding,
  MNN = mnn_embedding,
  Seurat_CCA = cca_result$embedding
)

embeddings <- imap(embeddings, ~ align_embedding(.x, metadata, .y))

embedding_dimensions <- tibble(
  method = names(embeddings),
  n_cells = vapply(embeddings, nrow, numeric(1)),
  n_dimensions = vapply(embeddings, ncol, numeric(1))
)

embedding_dimensions
```

# Technical Batch-Shift Metrics

For each method, cell type, and batch pair:

```text
R_tech = median cross-batch same-cell-type distance /
         median within-batch same-cell-type baseline
```

The target is `R_tech` near 1. The score `abs(log(R_tech))` is lower when the
ratio is closer to 1.

```{r compute-technical-metrics}
technical_results <- imap(embeddings, function(embedding, method_name) {
  compute_technical_metrics_for_embedding(
    embedding = embedding,
    metadata = metadata,
    batch_col = batch_col,
    cell_type_col = cell_type_col,
    method = method_name,
    min_cells_per_group = min_cells_per_group,
    max_cells_per_group = max_cells_per_group,
    max_pairs_per_comparison = max_pairs_per_comparison,
    seed = seed
  )
})

technical_summary <- bind_rows(map(technical_results, "summary"))
technical_distance_samples <- bind_rows(map(technical_results, "distances"))

readr::write_csv(
  technical_summary,
  file.path(tables_dir, "technical_summary.csv")
)
readr::write_csv(
  technical_distance_samples,
  file.path(tables_dir, "technical_distance_samples.csv")
)

technical_summary
```

```{r plot-technical-ratio-heatmap, fig.width=12, fig.height=8}
p_technical_ratio <- plot_technical_ratio_heatmap(technical_summary)
p_technical_ratio
save_plot(p_technical_ratio, "technical_ratio_heatmap.png", width = 12, height = 8)
```

```{r plot-technical-deviation-heatmap, fig.width=12, fig.height=8}
p_technical_deviation <- plot_technical_deviation_heatmap(technical_summary)
p_technical_deviation
save_plot(
  p_technical_deviation,
  "technical_deviation_heatmap.png",
  width = 12,
  height = 8
)
```

```{r plot-technical-scatter, fig.width=10, fig.height=7}
technical_scatter <- make_technical_scatter_data(technical_summary)
readr::write_csv(
  technical_scatter,
  file.path(tables_dir, "technical_before_after_scatter_data.csv")
)

p_technical_scatter <- plot_technical_scatter(technical_scatter)
p_technical_scatter
save_plot(p_technical_scatter, "technical_before_after_scatter.png", width = 10, height = 7)
```

```{r plot-cross-vs-within-overlay, fig.width=24, fig.height=28, dpi=300, out.width="100%"}
technical_overlay <- filter_overlay_distances(
  technical_distance_samples,
  selected_batch_pairs = params$selected_overlay_batch_pairs,
  selected_cell_types = params$selected_overlay_cell_types
)

p_cross_within <- plot_cross_vs_within_overlay(technical_overlay)
p_cross_within
save_plot(p_cross_within, "cross_vs_within_normalized_overlay.png", width = 24, height = 28)
```

# Biological Structure Preservation Metrics

For each method, batch, and target cell type, distances are centered on the
reference cell type. With `reference_cell_type = "alpha"`:

```text
relative_distance = pairwise alpha-to-target distance /
                    median alpha-to-alpha within-batch distance
```

The summary is the mean of pairwise relative distances, not a ratio of medians.

```{r compute-biological-metrics}
biological_results <- imap(embeddings, function(embedding, method_name) {
  compute_biological_metrics_for_embedding(
    embedding = embedding,
    metadata = metadata,
    batch_col = batch_col,
    cell_type_col = cell_type_col,
    method = method_name,
    reference_cell_type = reference_cell_type,
    min_cells_per_group = min_cells_per_group,
    max_cells_per_group = max_cells_per_group,
    max_pairs_per_comparison = max_pairs_per_comparison,
    seed = seed
  )
})

biological_summary <- bind_rows(map(biological_results, "summary"))
biological_distance_samples <- bind_rows(map(biological_results, "distances"))
biological_delta <- compute_biological_geometry_delta(biological_summary)

readr::write_csv(
  biological_summary,
  file.path(tables_dir, "biological_summary.csv")
)
readr::write_csv(
  biological_distance_samples,
  file.path(tables_dir, "biological_distance_samples.csv")
)
readr::write_csv(
  biological_delta,
  file.path(tables_dir, "biological_signed_geometry_delta.csv")
)

biological_summary
```

```{r plot-alpha-beta-distribution, fig.width=11, fig.height=7}
p_alpha_beta <- plot_alpha_beta_distribution(
  biological_distance_samples,
  beta_cell_type = beta_cell_type
)
p_alpha_beta
save_plot(p_alpha_beta, "alpha_beta_relative_distance_distribution.png", width = 11, height = 7)
```

```{r plot-mean-relative-heatmap, fig.width=12, fig.height=8}
p_mean_relative <- plot_mean_relative_heatmap(biological_summary)
p_mean_relative
save_plot(p_mean_relative, "mean_relative_separation_heatmap.png", width = 12, height = 8)
```

```{r plot-biological-scatter, fig.width=10, fig.height=7}
biological_scatter <- make_biological_scatter_data(biological_summary)
readr::write_csv(
  biological_scatter,
  file.path(tables_dir, "biological_before_after_scatter_data.csv")
)

p_biological_scatter <- plot_biological_scatter(biological_scatter)
p_biological_scatter
save_plot(p_biological_scatter, "biological_preservation_scatter.png", width = 10, height = 7)
```

```{r plot-signed-delta-heatmap, fig.width=12, fig.height=8}
p_signed_delta <- plot_signed_delta_heatmap(biological_delta)
p_signed_delta
save_plot(p_signed_delta, "signed_biological_geometry_delta_heatmap.png", width = 12, height = 8)
```

# Final Method Summary

```{r final-summary-table}
final_method_summary <- make_final_method_summary(
  technical_summary = technical_summary,
  biological_summary = biological_summary,
  biological_delta = biological_delta,
  beta_cell_type = beta_cell_type,
  baseline_method = "Uncorrected_PCA"
)

readr::write_csv(
  final_method_summary,
  file.path(tables_dir, "final_method_summary.csv")
)

final_method_summary
```

# Interpretation Guide

## Technical Ratio Heatmap

`R_tech` compares cross-batch same-cell-type distances with same-cell-type
within-batch distances in the same representation. Values near 1 indicate good
mixing. Values much greater than 1 suggest undercorrection. Values much less
than 1 suggest possible overcompression or overcorrection.

## Technical Deviation Heatmap

This shows `abs(log(R_tech))`. Lower is better because it penalizes ratios both
above and below 1.

## Before-vs-after Technical Scatter

Points below the `y = x` line usually indicate reduced batch shift compared with
uncorrected PCA. Points near the horizontal `y = 1` line are closest to the
target technical mixing level. Points far below 1 can be suspicious because they
may indicate artificial compression.

## Cross-vs-within Overlay

Cross-batch and within-batch distances are normalized by the same within-batch
baseline for that method, cell type, and batch pair. Good technical correction
means the cross-batch distribution overlaps the within-batch distribution.

## Alpha-to-beta Relative Distribution

For the beta target, values mostly above 1 mean beta cells remain separated from
the alpha baseline. If a corrected method pulls beta values toward 1, inspect it
for biological collapse.

## Mean Relative Separation Heatmap

Alpha should be near its own baseline. Beta and other distinct cell types should
usually remain greater than 1. Missing tiles are expected when a cell type is
absent or below the minimum cell-count threshold in a batch.

## Biological Preservation Scatter

The x-axis is the uncorrected PCA relative separation, and the y-axis is the
corrected relative separation. Movement toward 1 can indicate loss of biological
geometry, especially for alpha versus beta.

## Signed Biological Geometry Difference Heatmap

Values near 0 mean two batches agree about alpha-centered biological geometry.
Interpret this together with the alpha-beta collapse plot: batch agreement is
not useful if all methods agree because biology has been collapsed.

## Final Method Summary

A useful method has a low technical score, a low biological preservation score,
a low biological batch-agreement score, and an alpha-beta collapse risk safely
above 1. A method with good technical score but alpha-beta collapse risk close
to 1 may be overcorrecting.
