---
title: "Pairwise Batch Distance Analysis"
author: "Taher Alabbar"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
  pdf_document:
    toc: true
params:
  data_root: "data/panc8"
  batches:
    - "celseq"
    - "celseq2"
    - "fluidigmc1"
    - "smartseq2"
    - "indrop"
  primary_batches:
    - "celseq"
    - "celseq2"
  assay: "RNA"
  distance_layer: "scale.data"
  feature_mode: "variable"
  n_variable_features: 2000
  min_cells: 3
  min_features: 200
  min_cells_per_type: 20
  max_cells_per_type_per_batch: 250
  sample_per_group: 5000
  seed: 123
  write_outputs: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 7,
  dpi = 300,
  out.width = "100%"
)

library(Seurat)
library(Matrix)
library(dplyr)
library(ggplot2)

source(file.path("scripts", "pairwise_distance_utils.R"))

theme_set(theme_minimal(base_size = 11))
```

# Overview

This report compares cells with the same annotated cell type across pairs of
single-cell RNA-seq batches. For each batch pair and shared cell type, it
calculates all cross-batch Euclidean distances in the selected expression space.

The main plots show:

- one selected batch-pair comparison in detail
- all batch pairs overlaid within each cell type
- a batch-pair grid where each panel contains the cell-type distance distributions
- heatmaps of mean, median, and standard-deviation distance summaries

# Analysis Settings

```{r analysis-settings}
set.seed(params$seed)

batches <- as.character(unlist(params$batches, use.names = FALSE))
primary_batches <- as.character(unlist(params$primary_batches, use.names = FALSE))

if (length(primary_batches) != 2) {
  stop("primary_batches must contain exactly two batch names.")
}

if (!all(primary_batches %in% batches)) {
  stop("primary_batches must be included in params$batches.")
}

data_root <- resolve_data_root(params$data_root)
dataset_dirs <- make_dataset_dirs(data_root, batches)
primary_batch_label <- make_batch_label(primary_batches)

data.frame(
  setting = c(
    "data_root",
    "batches",
    "primary_batches",
    "distance_layer",
    "feature_mode",
    "min_cells_per_type",
    "max_cells_per_type_per_batch"
  ),
  value = c(
    data_root,
    paste(batches, collapse = ", "),
    paste(primary_batches, collapse = ", "),
    params$distance_layer,
    params$feature_mode,
    params$min_cells_per_type,
    params$max_cells_per_type_per_batch
  )
)
```

# Load And Preprocess Data

```{r load-data}
seu <- load_10x_batches(
  dataset_dirs = dataset_dirs,
  min_cells = params$min_cells,
  min_features = params$min_features
)

seu
```

```{r preprocess-data}
seu <- NormalizeData(seu, verbose = FALSE)
seu <- FindVariableFeatures(
  seu,
  selection.method = "vst",
  nfeatures = params$n_variable_features,
  verbose = FALSE
)

if (identical(params$distance_layer, "scale.data")) {
  scale_features <- if (identical(params$feature_mode, "all")) {
    rownames(seu)
  } else {
    VariableFeatures(seu)
  }

  seu <- ScaleData(seu, features = scale_features, verbose = FALSE)
}
```

```{r cell-counts}
table(seu$batch)
table(seu$cell_type, seu$batch, useNA = "ifany")
```

```{r select-features}
expression_matrix <- get_expression_layer(
  seurat_object = seu,
  assay = params$assay,
  layer = params$distance_layer
)

distance_features <- select_distance_features(
  seurat_object = seu,
  expression_matrix = expression_matrix,
  feature_mode = params$feature_mode,
  n_features = params$n_variable_features
)

data.frame(
  distance_layer = params$distance_layer,
  feature_mode = params$feature_mode,
  n_features = length(distance_features)
)
```

# Primary Batch Pair

```{r primary-distances}
primary_distances <- calculate_matching_celltype_distances(
  seurat_object = seu,
  batch_a = primary_batches[1],
  batch_b = primary_batches[2],
  assay = params$assay,
  layer = params$distance_layer,
  features = distance_features,
  min_cells_per_type = params$min_cells_per_type,
  max_cells_per_type_per_batch = params$max_cells_per_type_per_batch,
  seed = params$seed
)

primary_summary <- summarise_distances(
  primary_distances,
  group_cols = "cell_type"
) |>
  arrange(mean_distance)

primary_summary
```

```{r primary-cell-types}
sort(unique(primary_distances$cell_type))
```

```{r primary-histograms}
plot_celltype_histograms(
  primary_distances,
  title = "Cross-batch distances for matching cell types",
  subtitle = paste(primary_batch_label, "in", params$distance_layer)
)
```

```{r primary-boxplot, fig.height=6}
plot_celltype_boxplot(
  primary_distances,
  title = "Distance distributions for matching cell types",
  subtitle = paste(primary_batch_label, "in", params$distance_layer)
)
```

```{r primary-density, fig.height=6}
plot_celltype_density(
  primary_distances,
  title = "Distance distributions for matching cell types",
  subtitle = paste(primary_batch_label, "in", params$distance_layer)
)
```

# All Batch Pairs

```{r all-pair-distances}
all_matching_distances <- calculate_all_batchpair_celltype_distances(
  seurat_object = seu,
  batches = batches,
  assay = params$assay,
  layer = params$distance_layer,
  features = distance_features,
  min_cells_per_type = params$min_cells_per_type,
  max_cells_per_type_per_batch = params$max_cells_per_type_per_batch,
  seed = params$seed
)

dim(all_matching_distances)
```

```{r all-pair-comparisons}
all_matching_distances |>
  distinct(batch_pair, cell_type, n_cells_a, n_cells_b) |>
  arrange(batch_pair, cell_type)
```

```{r plot-sample}
all_matching_distances_sample <- sample_distance_table(
  distance_table = all_matching_distances,
  group_cols = c("batch_pair", "cell_type"),
  n_per_group = params$sample_per_group,
  seed = params$seed
)
```

```{r all-pair-density-by-celltype, fig.width=16, fig.height=12}
plot_batchpair_density_by_celltype(
  all_matching_distances_sample,
  title = "Cross-batch same-cell-type distance distributions",
  subtitle = "Each panel is one cell type; colors show batch pairs"
)
```

```{r all-pair-matrix-density, fig.width=16, fig.height=11}
plot_batchpair_matrix_density(
  all_matching_distances_sample,
  title = "Batch-pair grid of same-cell-type distance distributions",
  subtitle = "Each grid panel is one batch pair; colors show cell types"
)
```

# Summary Heatmaps

```{r all-pair-summary}
all_distance_summary <- summarise_distances(
  all_matching_distances,
  group_cols = c("batch_pair", "cell_type")
) |>
  arrange(cell_type, mean_distance)

all_distance_summary
```

```{r median-distance-heatmap, fig.width=11, fig.height=7}
plot_distance_heatmap(
  all_distance_summary,
  value_column = "median_distance",
  title = "Median distances between matching cell types across batches",
  subtitle = paste("Euclidean distances in", params$distance_layer),
  fill_label = "Median\ndistance"
)
```

```{r sd-distance-heatmap, fig.width=11, fig.height=7}
plot_distance_heatmap(
  all_distance_summary,
  value_column = "sd_distance",
  title = "Standard deviation of matching-cell-type distances",
  subtitle = paste("Euclidean distances in", params$distance_layer),
  fill_label = "SD\ndistance"
)
```

```{r mean-distance-heatmap, fig.width=11, fig.height=7}
plot_distance_heatmap(
  all_distance_summary,
  value_column = "mean_distance",
  title = "Mean distances between matching cell types across batches",
  subtitle = paste("Euclidean distances in", params$distance_layer),
  fill_label = "Mean\ndistance"
)
```

# Export Tables

```{r export-tables}
if (isTRUE(params$write_outputs)) {
  dir.create("output", showWarnings = FALSE)

  write.csv(
    primary_summary,
    file = file.path("output", "primary_pair_distance_summary.csv"),
    row.names = FALSE
  )

  write.csv(
    all_distance_summary,
    file = file.path("output", "all_pair_distance_summary.csv"),
    row.names = FALSE
  )
}
```

# Session Info

```{r session-info}
sessionInfo()
```
