---
title: "Single-Cell Batch-Correction Evaluation"
author: "Taher Alabbar"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
  pdf_document:
    toc: true
params:
  seurat_rds: !r NULL
  seurat_object_name: !r NULL
  tenx_data_root: "data/panc8"
  tenx_batches: !r c("celseq", "celseq2", "fluidigmc1", "smartseq2", "indrop")
  batch_col: "batch"
  cell_type_col: "cell_type"
  assay: "RNA"
  reference_cell_type: "alpha"
  beta_cell_type: "beta"
  n_dims: 30
  n_variable_features: 2000
  scale_all_genes: true
  future_globals_max_size_gb: 8
  min_cells_per_group: 10
  max_cells_per_group: 500
  max_pairs_per_comparison: 10000
  seed: 123
  output_dir: "output"
  selected_overlay_batch_pairs: !r NULL
  selected_overlay_cell_types: !r c("alpha", "beta")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.align = "center",
  dpi = 300,
  out.width = "100%"
)

source(file.path("scripts", "package_setup.R"))
source(file.path("scripts", "data_loading.R"))
source(file.path("scripts", "evaluation_utils.R"))
source(file.path("scripts", "plot_utils.R"))
```

# Overview

This report evaluates batch-correction methods for annotated single-cell RNA-seq
data. It compares:

- `Uncorrected_PCA`
- `FastMNN`
- `MNN`
- `Seurat_CCA`

Each method is evaluated with two complementary diagnostics:

- Technical mixing: same-cell-type cross-batch distances should resemble
  same-cell-type within-batch distances.
- Biological preservation: alpha-centered cell-type separation should remain
  similar to the uncorrected representation.

# Configuration

```{r config}
batch_col <- params$batch_col
cell_type_col <- params$cell_type_col
assay_name <- params$assay
reference_cell_type <- params$reference_cell_type
beta_cell_type <- params$beta_cell_type
n_dims <- params$n_dims
n_variable_features <- params$n_variable_features
scale_all_genes <- params$scale_all_genes
future_globals_max_size_gb <- params$future_globals_max_size_gb
min_cells_per_group <- params$min_cells_per_group
max_cells_per_group <- params$max_cells_per_group
max_pairs_per_comparison <- params$max_pairs_per_comparison
seed <- params$seed

options(future.globals.maxSize = future_globals_max_size_gb * 1024^3)
future::plan(future::sequential)
set.seed(seed)

output_dir <- params$output_dir
figures_dir <- file.path(output_dir, "figures")
tables_dir <- file.path(output_dir, "tables")
dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tables_dir, recursive = TRUE, showWarnings = FALSE)

tibble(
  setting = c(
    "tenx_data_root",
    "tenx_batches",
    "batch_col",
    "cell_type_col",
    "assay",
    "reference_cell_type",
    "n_dims",
    "max_cells_per_group",
    "max_pairs_per_comparison"
  ),
  value = c(
    params$tenx_data_root,
    paste(params$tenx_batches, collapse = ", "),
    batch_col,
    cell_type_col,
    assay_name,
    reference_cell_type,
    n_dims,
    max_cells_per_group,
    max_pairs_per_comparison
  )
)
```

# Input Seurat Object

Use one of these inputs:

- `seurat_rds`: path to a saved `.rds` Seurat object.
- `seurat_object_name`: name of an already-loaded Seurat object.
- `tenx_data_root`: folder containing 10x batch directories.

```{r load-seurat-object}
seu <- load_input_seurat_object(
  seurat_rds_path = params$seurat_rds,
  seurat_object_name = params$seurat_object_name,
  tenx_data_root = params$tenx_data_root,
  tenx_batches = params$tenx_batches,
  batch_col = batch_col,
  cell_type_col = cell_type_col
)

if (!inherits(seu, "Seurat")) {
  stop("`seu` must be a Seurat object.", call. = FALSE)
}

seu
```

# Normalize And Select Variable Genes

```{r qc-normalize}
SeuratObject::DefaultAssay(seu) <- assay_name

seu <- Seurat::NormalizeData(seu, verbose = FALSE)
seu <- Seurat::FindVariableFeatures(
  seu,
  selection.method = "vst",
  nfeatures = n_variable_features,
  verbose = FALSE
)

pca_features <- SeuratObject::VariableFeatures(seu)

if (length(pca_features) == 0) {
  stop("No variable features were selected for PCA.", call. = FALSE)
}

features_to_scale <- if (isTRUE(scale_all_genes)) {
  rownames(seu)
} else {
  pca_features
}

seu <- Seurat::ScaleData(
  seu,
  features = features_to_scale,
  verbose = FALSE
)

tibble(
  n_cells = ncol(seu),
  n_genes = nrow(seu),
  n_variable_features = length(pca_features),
  n_scaled_features = length(features_to_scale)
)
```

```{r cell-counts}
table(seu[[batch_col]][, 1])
table(seu[[cell_type_col]][, 1], seu[[batch_col]][, 1], useNA = "ifany")
```

# Build Representations

All distance comparisons are made inside a representation and normalized inside
that same representation. The report does not compare raw original-space
distances directly against corrected embedding distances.

```{r build-representations}
metadata <- check_required_metadata(seu, batch_col, cell_type_col, assay_name)

pca_result <- run_uncorrected_pca_embedding(
  seu = seu,
  assay_name = assay_name,
  n_dims = n_dims,
  n_variable_features = n_variable_features,
  seed = seed
)

metadata <- check_required_metadata(pca_result$object, batch_col, cell_type_col, assay_name)
metadata <- metadata[colnames(pca_result$object), , drop = FALSE]

fastmnn_embedding <- run_fastmnn_embedding(
  seu = pca_result$object,
  metadata = metadata,
  batch_col = batch_col,
  assay_name = assay_name,
  features = pca_result$features,
  n_dims = n_dims,
  seed = seed + 1
)

mnn_embedding <- run_mnn_embedding(
  seu = pca_result$object,
  metadata = metadata,
  batch_col = batch_col,
  assay_name = assay_name,
  features = pca_result$features,
  n_dims = n_dims,
  seed = seed + 2
)

cca_result <- run_seurat_cca_embedding(
  seu = seu,
  batch_col = batch_col,
  assay_name = assay_name,
  n_dims = n_dims,
  n_variable_features = n_variable_features,
  seed = seed + 3
)

embeddings <- list(
  Uncorrected_PCA = pca_result$embedding,
  FastMNN = fastmnn_embedding,
  MNN = mnn_embedding,
  Seurat_CCA = cca_result$embedding
)

embeddings <- imap(embeddings, ~ align_embedding(.x, metadata, .y))

embedding_dimensions <- tibble(
  method = names(embeddings),
  n_cells = vapply(embeddings, nrow, numeric(1)),
  n_dimensions = vapply(embeddings, ncol, numeric(1))
)

embedding_dimensions
```

# Technical Batch-Shift Metrics

For each method, cell type, and batch pair:

```text
R_tech = median cross-batch same-cell-type distance /
         median within-batch same-cell-type baseline
```

The target is `R_tech` near 1. The score `abs(log(R_tech))` is lower when the
ratio is closer to 1.

```{r compute-technical-metrics}
technical_results <- imap(embeddings, function(embedding, method_name) {
  compute_technical_metrics_for_embedding(
    embedding = embedding,
    metadata = metadata,
    batch_col = batch_col,
    cell_type_col = cell_type_col,
    method = method_name,
    min_cells_per_group = min_cells_per_group,
    max_cells_per_group = max_cells_per_group,
    max_pairs_per_comparison = max_pairs_per_comparison,
    seed = seed
  )
})

technical_summary <- bind_rows(map(technical_results, "summary"))
technical_distance_samples <- bind_rows(map(technical_results, "distances"))

readr::write_csv(technical_summary, file.path(tables_dir, "technical_summary.csv"))
readr::write_csv(technical_distance_samples, file.path(tables_dir, "technical_distance_samples.csv"))

technical_summary
```

```{r plot-technical-ratio-heatmap, fig.width=12, fig.height=8}
p_technical_ratio <- plot_technical_ratio_heatmap(technical_summary)
p_technical_ratio
save_plot(p_technical_ratio, "technical_ratio_heatmap.png", width = 12, height = 8)
```

```{r plot-technical-deviation-heatmap, fig.width=12, fig.height=8}
p_technical_deviation <- plot_technical_deviation_heatmap(technical_summary)
p_technical_deviation
save_plot(p_technical_deviation, "technical_deviation_heatmap.png", width = 12, height = 8)
```

```{r plot-technical-scatter, fig.width=10, fig.height=7}
technical_scatter <- make_technical_scatter_data(technical_summary)
readr::write_csv(technical_scatter, file.path(tables_dir, "technical_before_after_scatter_data.csv"))

p_technical_scatter <- plot_technical_scatter(technical_scatter)
p_technical_scatter
save_plot(p_technical_scatter, "technical_before_after_scatter.png", width = 10, height = 7)
```

```{r plot-cross-vs-within-overlay, fig.width=24, fig.height=28}
technical_overlay <- filter_overlay_distances(
  technical_distance_samples,
  selected_batch_pairs = params$selected_overlay_batch_pairs,
  selected_cell_types = params$selected_overlay_cell_types
)

p_cross_within <- plot_cross_vs_within_overlay(technical_overlay)
p_cross_within
save_plot(p_cross_within, "cross_vs_within_normalized_overlay.png", width = 24, height = 28)
```

# Biological Structure Preservation Metrics

For each method, batch, and target cell type, distances are centered on the
reference cell type. With `reference_cell_type = "alpha"`:

```text
relative_distance = pairwise alpha-to-target distance /
                    median alpha-to-alpha within-batch distance
```

The summary is the mean of pairwise relative distances, not a ratio of medians.

```{r compute-biological-metrics}
biological_results <- imap(embeddings, function(embedding, method_name) {
  compute_biological_metrics_for_embedding(
    embedding = embedding,
    metadata = metadata,
    batch_col = batch_col,
    cell_type_col = cell_type_col,
    method = method_name,
    reference_cell_type = reference_cell_type,
    min_cells_per_group = min_cells_per_group,
    max_cells_per_group = max_cells_per_group,
    max_pairs_per_comparison = max_pairs_per_comparison,
    seed = seed
  )
})

biological_summary <- bind_rows(map(biological_results, "summary"))
biological_distance_samples <- bind_rows(map(biological_results, "distances"))
biological_delta <- compute_biological_geometry_delta(biological_summary)

readr::write_csv(biological_summary, file.path(tables_dir, "biological_summary.csv"))
readr::write_csv(biological_distance_samples, file.path(tables_dir, "biological_distance_samples.csv"))
readr::write_csv(biological_delta, file.path(tables_dir, "biological_signed_geometry_delta.csv"))

biological_summary
```

```{r plot-alpha-beta-distribution, fig.width=11, fig.height=7}
p_alpha_beta <- plot_alpha_beta_distribution(
  biological_distance_samples,
  beta_cell_type = beta_cell_type
)
p_alpha_beta
save_plot(p_alpha_beta, "alpha_beta_relative_distance_distribution.png", width = 11, height = 7)
```

```{r plot-mean-relative-heatmap, fig.width=12, fig.height=8}
p_mean_relative <- plot_mean_relative_heatmap(biological_summary)
p_mean_relative
save_plot(p_mean_relative, "mean_relative_separation_heatmap.png", width = 12, height = 8)
```

```{r plot-biological-scatter, fig.width=10, fig.height=7}
biological_scatter <- make_biological_scatter_data(biological_summary)
readr::write_csv(biological_scatter, file.path(tables_dir, "biological_before_after_scatter_data.csv"))

p_biological_scatter <- plot_biological_scatter(biological_scatter)
p_biological_scatter
save_plot(p_biological_scatter, "biological_preservation_scatter.png", width = 10, height = 7)
```

```{r plot-signed-delta-heatmap, fig.width=12, fig.height=8}
p_signed_delta <- plot_signed_delta_heatmap(biological_delta)
p_signed_delta
save_plot(p_signed_delta, "signed_biological_geometry_delta_heatmap.png", width = 12, height = 8)
```

# Final Method Summary

```{r final-summary-table}
final_method_summary <- make_final_method_summary(
  technical_summary = technical_summary,
  biological_summary = biological_summary,
  biological_delta = biological_delta,
  beta_cell_type = beta_cell_type,
  baseline_method = "Uncorrected_PCA"
)

readr::write_csv(final_method_summary, file.path(tables_dir, "final_method_summary.csv"))

final_method_summary
```

# Interpretation Guide

Lower technical score means corrected same-cell-type cells mix better across
batches. Biological preservation should be read alongside the technical score:
a method that mixes batches well is only useful if alpha-centered cell-type
separation does not collapse toward the reference baseline.

# Session Info

```{r session-info}
sessionInfo()
```
